Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-03 _tex/index.tex
Built site for gh-pages

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
be suitable for applications beyond the psycho-physical contexts for
which it was created. Thurstone himself cautioned that its use ``should
not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}, acknowledging that some assumptions
could prove problematic in the presence of complex traits or
heterogeneous stimuli, such as handwriting or English compositions
\citep[pp.~374]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve these traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice: the zero correlation and equal
dispersion between stimuli.

As outlined in Section~\ref{sec-thurstone_theory}, the assumption of
\emph{zero correlation between stimuli} implies that, during a pairwise
comparison, a judge's perception of quality in one text does not
influence his perception of the same trait in another text. Thurstone
attributed this independence to the cancellation of potential judges'
biases, driven by two opposing and equally weighted effects occurring
during the pairwise comparisons \citep[pp.~268]{Thurstone_1927b}.
\citet{Andrich_1978} mathematically demonstrated this cancellation using
the BTL model under the assumption of discriminal processes with
additive biases. However, it is easy to imagine at least two scenarios
where the zero correlation assumption almost certainly does not hold:
when the pairwise comparison involves multidimensional, complex traits
with heterogeneous stimuli and when an additional hierarchical structure
is relevant to the stimuli.

@@

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures, the
statistical handling of this extra source of dependency between the
stimuli has been inadequate. For example, when CJ data includes multiple
samples of stimuli from the same individuals, researchers often rely on
(average) estimated BTL scores to conduct subsequent analyses and tests
at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the BTL scores, which generates additional
statistical and measurement issues, as discussed in section
Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature emphasizes the
importance of addressing factors that create dependencies between
stimuli, as failing to do so can lead to inaccurate conclusions about a
trait's reliability and, by extension, its validity. For instance,
neglecting additional traits relevant to the stimuli, such as judges'
biases, often leads to a dimensional mismatch in the model. This
mismatch can potentially inflate the reliability of the trait
\citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias into the
trait's estimates \citep{Ackerman_1989}. Similarly, failing to account
for hierarchical (grouping) structures reduces the precision of model
parameter estimates, further amplifying the overestimation of
reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
undermine the trait's reliability, and because ``validity cannot exist
without reliability'' \citep[pp.~2]{Perron_et_al_2015}, they ultimately
compromise the validity of results derived from the data modeling.

Fortunately, the same literature offers solutions for addressing the
factors that create these dependencies between stimuli.
\citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend integrating
judges' biases into the BTL model. Furthermore, the literature advocates
for incorporating relevant hierarchical structures into the data model
to account for these dependencies. Taken together, these additions can
result in a model that combines a Multilevel Structural Equation Model
(MSEM) \citep[chap.~26]{Hoyle_et_al_2023} with a multidimensional or
two-parameter logistic IRT model \citep[chap.~15]{Hoyle_et_al_2023},
depending on the theoretical assumptions a researcher makes about the
judges' biases.

>>>>

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. As a result, its assumptions may not
be suitable for applications beyond the psycho-physical contexts for
which it was created. Thurstone himself cautioned that its use ``should
not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}, acknowledging that some assumptions
could prove problematic in the presence of complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}. Consequently,
given that modern CJ applications frequently involve these traits and
stimuli, two main assumptions of Case V and, by extension, of the BTL
model may not consistently hold in theory or practice: the equal
dispersion and zero correlation between stimuli.

On the one hand, the assumption of \emph{equal dispersion between
stimuli} asserts that the variability in perceived trait levels remains
consistent across all stimuli. However, Thurstone noted that this
assumption may fail when researchers assess complex traits or
heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}, as these can
introduce greater discrepancies into judgments due to their unique
characteristics. One example of this is the evaluation of creativity. In
this context, for instance, it is plausible that a student completes an
assignment in a highly creative yet unexpected way. Judges, unprepared
for such originality, may experience greater discrepancies in their
judgments of the assignment. As a result, the discriminal difference
associated with this assignment is expected to show a broader
dispersion, reflecting the larger variation in its discriminal process
compared to others. Although direct evidence for this specific scenario
is lacking, {studies such as \citet[chap.~8]{vanDaal_2020} suggest that
misfitting texts tend to exhibit larger dispersion, likely due to
increased confusion among judges regarding the texts.}

{what are the challenges when the equal dispersion assumption is
violated?}

{how do we mitigate these risks?}

On the other hand hand, the assumption of \emph{zero correlation between
stimuli} implies that, during a pairwise comparison, a judge's
perception of quality in one text does not influence his perception of
the same trait in another text (see Section~\ref{sec-thurstone_theory}).
Thurstone attributed this independence to the cancellation of potential
judges' biases, driven by two opposing and equally weighted effects
occurring during the pairwise comparisons
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} mathematically
demonstrated this cancellation using the BTL model under the assumption
of discriminal processes with additive biases. However, it is easy to
imagine at least two scenarios where the zero correlation assumption
almost certainly does not hold: when the pairwise comparison involves
multidimensional, complex traits with heterogeneous stimuli and when an
additional hierarchical structure is relevant to the stimuli.

@@

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures, the
statistical handling of this extra source of dependency between stimuli
has been inadequate. For example, when CJ data includes multiple samples
of stimuli from the same individuals, researchers often rely on
(average) estimated BTL scores to conduct subsequent analyses and tests
at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the BTL scores, which generates additional
statistical and measurement issues, as discussed in section
Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature emphasizes the
importance of addressing the factors that create dependencies between
stimuli, as failing to do so can lead to inaccurate conclusions about a
trait's reliability and, by extension, its validity. For instance,
neglecting additional traits relevant to the stimuli, such as judges'
biases, often leads to a dimensional mismatch in the statistical model
used for analysis. This mismatch can potentially inflate the reliability
of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, worse, introduce bias
into the trait's estimates \citep{Ackerman_1989}. Similarly, failing to
account for hierarchical (grouping) structures reduces the precision of
model parameter estimates, further amplifying the overestimation of
reliability \citep[pp.~482]{Hoyle_et_al_2023}. These issues collectively
undermine the trait's reliability and ultimately compromise the validity
of the trait's estimates \citep[pp.~2]{Perron_et_al_2015}.

Fortunately, the same literature offers solutions for addressing these
issues. \citet{Andrich_1978} and \citet{Wainer_et_al_1978} recommend
integrating judges' biases into the BTL model. Moreover, the literature
advocates for the incorporation of relevant hierarchical structures into
the statistical model to account for these dependencies. Together, these
additions can result in a model resembling a Multilevel Structural
Equation Model (MSEM) \citep[chap.~26]{Hoyle_et_al_2023} combined with a
multidimensional or two-parameter logistic IRT model
\citep[chap.~15]{Hoyle_et_al_2023}, depending on the theoretical and
statistical treatment of judges' biases.
